---
title: "Getting Started with MLfused"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with MLfused}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

MLfused fits multinomial logistic regression models that incorporate external
machine learning predictions via empirical-likelihood (EL) constraints.
The key idea is straightforward: you have a primary dataset with an outcome
of K classes and p covariates, plus external ML predictions (possibly from a
larger or differently-distributed dataset) that provide grouped class
probabilities. MLfused fuses these two sources of information to produce
coefficient estimates with smaller variance than a standard multinomial model
fit on the primary data alone.

This vignette walks through a minimal example: generating synthetic data,
fitting a fused model, and comparing the results with a baseline
multinomial logistic regression.

## Generate synthetic data

We create a multinomial outcome with K = 3 classes, p = 3 covariates, and
n = 500 primary observations. Class 1 is the baseline (reference) category.

```{r generate-data}
library(MLfused)
set.seed(42)

n <- 500
p <- 3
K <- 3

# Covariates: standard normal
X <- matrix(rnorm(n * p), nrow = n, ncol = p)
colnames(X) <- paste0("x", 1:p)

# True coefficients: (K-1) intercepts and p x (K-1) slopes
beta_true  <- c(0.3, -0.2)
Theta_true <- matrix(c(
   1.0, -0.5,
  -0.8,  0.6,
   0.4,  0.3
), nrow = p, ncol = K - 1, byrow = TRUE)

# Generate class probabilities via softmax and draw outcomes
eta <- cbind(0, sweep(X %*% Theta_true, 2, beta_true, "+"))
P_true <- exp(eta) / rowSums(exp(eta))
y <- apply(P_true, 1, function(prob) sample(1:K, 1, prob = prob))

table(y)
```

## Create synthetic external predictions

In practice, `qhat` comes from an external ML model (e.g., XGBoost trained
on a larger auxiliary dataset). Here we simulate it by adding noise to the
true grouped probabilities.

We define two groups: group 1 = {class 1} (reference) and group 2 =
{class 2, class 3}. The external model predicts the probability of the
non-reference group (group 2), so `qhat` is an n x 1 matrix.

```{r external-predictions}
# Define groups: group 1 = {class 1}, group 2 = {classes 2, 3}
groups <- list(c(1), c(2, 3))

# True group-2 probability
q_true <- P_true[, 2] + P_true[, 3]

# Simulate external predictions with noise
qhat    <- matrix(pmin(pmax(q_true + rnorm(n, 0, 0.05), 0.01), 0.99), ncol = 1)
qhat_se <- matrix(rep(0.05, n), ncol = 1)
```

## Build the H constraint matrix

The H matrix encodes the basis functions for the EL constraint. With
`n.q = 0`, we get an intercept plus centered raw covariates. With `n.q = 1`,
we get an intercept plus natural spline bases with 1 internal knot per
covariate.

```{r build-H}
# Use the first two covariates as constraint features
phi_idx <- c(1, 2)

# n.q = 0: intercept + centered covariates
H_base <- build_H(X, phi_idx = phi_idx, n.q = 0)
dim(H_base)

# For L = 2 groups, Lm1 = 1, so no interleaving is needed
Hmat <- build_Hmat_interleaved(H_base, Lm1 = length(groups) - 1)
dim(Hmat)
```

## Set up initial parameters

We initialize the fused model at the multinomial MLE. The parameter vector
is packed in the order (beta, Theta, alpha, tmat).

```{r init-params}
# Initialize at zeros (or use nnet::multinom estimates if available)
beta0  <- rep(0, K - 1)
Theta0 <- matrix(0, nrow = p, ncol = K - 1)
alpha0 <- rep(0, K - 1)
Lm1    <- length(groups) - 1
H_per  <- ncol(H_base)
tmat0  <- matrix(0, nrow = Lm1, ncol = H_per)

# Pack into initial parameter vector
par0 <- pack.hard(beta0, Theta0, alpha0, tmat0)
```

## Fit the fused model

We use `ML_fused_fast()`, the optimized hard-mode solver. The key tuning
parameters are:

- `lambda`: regularization for the EL constraint (typically set to the
  external sample size)
- `learning.rate`: Newton step size
- `tau_tmat`: scaling for the t-parameter block

```{r fit-model}
fit <- ML_fused_fast(
  par           = par0,
  X             = X,
  y             = y,
  qhat          = qhat,
  qhat.se       = qhat_se,
  Hmat          = Hmat,
  groups        = groups,
  maxit         = 300,
  tol           = 1e-6,
  learning.rate = 0.1,
  lambda        = 5000,
  lambda.diag   = 0,
  compute_se    = TRUE,
  tau_tmat      = 0.01
)

cat("Convergence:", fit$conv, "\n")
cat("Best gradient norm:", format(fit$best_grad_norm, digits = 3), "\n")
```

## Extract results

The fitted object contains unpacked parameter estimates and multiple SE
variants. The score-based adjusted SE (`se.score.adj`) is recommended for
inference as it is consistent even under mild model misspecification and
accounts for uncertainty in the external predictions.

```{r extract-results}
# Coefficient estimates
cat("Intercepts (beta):\n")
print(round(fit$par$beta, 4))

cat("\nSlope matrix (Theta), p x (K-1):\n")
print(round(fit$par$Theta, 4))

# Score-based adjusted SEs
se <- fit$se.score.adj
cat("\nSE for beta:\n")
print(round(se$beta, 4))

cat("\nSE for Theta:\n")
print(round(se$Theta, 4))

# 95% confidence intervals for beta
ci_lo <- fit$par$beta - 1.96 * se$beta
ci_hi <- fit$par$beta + 1.96 * se$beta
cat("\n95% CI for beta:\n")
for (k in seq_along(fit$par$beta)) {
  cat(sprintf("  Class %d: [%.4f, %.4f]\n", k + 1, ci_lo[k], ci_hi[k]))
}
```

## Compare with baseline multinomial regression

```{r compare-nnet, eval = requireNamespace("nnet", quietly = TRUE)}
# Fit standard multinomial logistic regression
fit_mn <- nnet::multinom(y ~ X, trace = FALSE)
cf_mn  <- coef(fit_mn)
if (is.vector(cf_mn)) cf_mn <- matrix(cf_mn, nrow = 1)

beta_mn  <- cf_mn[, "(Intercept)"]
Theta_mn <- t(cf_mn[, -1, drop = FALSE])
se_mn    <- summary(fit_mn)$standard.errors

cat("=== Intercept comparison ===\n")
cat(sprintf("  True:     %s\n", paste(round(beta_true, 4), collapse = ", ")))
cat(sprintf("  Multinom: %s\n", paste(round(beta_mn, 4), collapse = ", ")))
cat(sprintf("  Fused:    %s\n", paste(round(fit$par$beta, 4), collapse = ", ")))

cat("\n=== Slope comparison (column = class, row = covariate) ===\n")
cat("True:\n"); print(round(Theta_true, 4))
cat("Multinom:\n"); print(round(Theta_mn, 4))
cat("Fused:\n"); print(round(fit$par$Theta, 4))

# Compare SEs for intercepts
se_mn_beta <- se_mn[, "(Intercept)"]
cat("\n=== SE for intercepts ===\n")
cat(sprintf("  Multinom: %s\n", paste(round(se_mn_beta, 4), collapse = ", ")))
cat(sprintf("  Fused:    %s\n", paste(round(se$beta, 4), collapse = ", ")))
```

## Visualize the comparison

```{r comparison-plot, fig.width = 8, fig.height = 5, eval = requireNamespace("nnet", quietly = TRUE)}
# Build a data frame for plotting
coef_names <- c("intercept", colnames(X))
results <- data.frame(
  coef   = rep(coef_names, times = K - 1),
  class  = rep(paste0("Class ", 2:K), each = p + 1),
  true   = c(rbind(beta_true, Theta_true)),
  fused  = c(rbind(fit$par$beta, fit$par$Theta)),
  fused_se = c(rbind(se$beta, se$Theta)),
  multinom = c(rbind(beta_mn, Theta_mn)),
  multinom_se = c(rbind(se_mn[, "(Intercept)"], t(se_mn[, -1, drop = FALSE])))
)

# Reshape for ggplot
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  df_long <- rbind(
    data.frame(
      coef     = results$coef,
      class    = results$class,
      method   = "Multinomial",
      estimate = results$multinom,
      se       = results$multinom_se,
      stringsAsFactors = FALSE
    ),
    data.frame(
      coef     = results$coef,
      class    = results$class,
      method   = "Fused",
      estimate = results$fused,
      se       = results$fused_se,
      stringsAsFactors = FALSE
    )
  )
  df_long$coef <- factor(df_long$coef, levels = coef_names)

  # True values for reference lines
  df_true <- data.frame(
    coef  = factor(rep(coef_names, times = K - 1), levels = coef_names),
    class = rep(paste0("Class ", 2:K), each = p + 1),
    true  = results$true
  )

  p_compare <- ggplot(df_long, aes(x = coef, y = estimate,
                                    color = method, shape = method)) +
    geom_hline(yintercept = 0, color = "gray80", linewidth = 0.4) +
    geom_hline(data = df_true, aes(yintercept = true),
               linetype = "dashed", color = "gray40", linewidth = 0.3) +
    geom_errorbar(aes(ymin = estimate - 1.96 * se,
                      ymax = estimate + 1.96 * se),
                  position = position_dodge(width = 0.5),
                  width = 0.25, linewidth = 0.4) +
    geom_point(size = 2.5, position = position_dodge(width = 0.5)) +
    facet_wrap(~ class) +
    scale_color_manual(values = c("Multinomial" = "steelblue",
                                  "Fused" = "tomato3")) +
    scale_shape_manual(values = c("Multinomial" = 16, "Fused" = 17)) +
    theme_bw(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "bottom") +
    labs(x = NULL, y = "Coefficient estimate",
         title = "Multinomial vs Fused: Coefficient Estimates",
         subtitle = "Error bars = 95% CI; dashed lines = true values",
         color = "Method", shape = "Method")

  print(p_compare)
}
```

The fused estimator typically shows smaller confidence intervals than the
standard multinomial model, especially when the external predictions are
informative and the primary sample is moderate in size.

## Next steps
- See `vignette("nhanes-application")` for a real-data example using NHANES
  blood pressure data.
- Increase `n.q` in `build_H()` to allow more flexible constraint functions
  via natural spline bases.
- Use `ML_fused()` for an alternative Newton solver with mini-batch and
  feasibility masking support.
