---
title: "Getting Started with MLfused"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with MLfused}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 8, fig.height = 5, dpi = 150)
has_deps <- requireNamespace("xgboost", quietly = TRUE) &&
            requireNamespace("nnet", quietly = TRUE) &&
            requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(eval = has_deps)
```

MLfused fits multinomial logistic regression that borrows strength from
external ML predictions via empirical-likelihood constraints. This demo
compares the standard **MLE** with the **Fused MLE (FMLE)** using the
same simulation design as the paper (K=3 classes, L=2 groups, no shift).

## 1. Setup and true parameters

```{r data}
library(MLfused)
library(xgboost)
library(ggplot2)

seed <- 42
n1 <- 500; n2 <- 10000; p <- 4; K <- 3
groups <- list(c(1, 2), c(3))
rho <- 0.8
nq  <- 1           # spline internal knots
phi_idx <- 1:4

Theta_true <- matrix(c(1, -1, -1, 1, 1, 1, -1, 1),
                     nrow = p, ncol = K - 1, byrow = TRUE)
beta_true  <- c(0.2, -0.1)
alpha_true <- c(0.35, -0.25)

Sigma <- matrix(rho, p, p); diag(Sigma) <- 1
```

## 2. Generate external data and train XGBoost

```{r xgboost}
set.seed(seed + 1)
X2   <- MASS::mvrnorm(n2, rep(0, p), Sigma)
eta2 <- X2 %*% Theta_true + matrix(alpha_true, n2, K - 1, byrow = TRUE)
P2   <- softmax_first(eta2)
G2   <- apply(P2, 1, function(pp) sample.int(K, 1, prob = pp))
W2   <- ifelse(G2 %in% c(1, 2), 1L, 2L)

dtrain  <- xgb.DMatrix(data = X2[, phi_idx], label = as.numeric(W2 == 1))
xgb_fit <- xgb.train(
  params  = list(objective = "binary:logistic", eval_metric = "logloss",
                 max_depth = 4, eta = 0.1, subsample = 0.5,
                 min_child_weight = 1, nthread = 1),
  data = dtrain, nrounds = 300, verbose = 0
)
```

## 3. Generate primary data and predict qhat

```{r primary}
set.seed(seed + n1 * 10000)
X1   <- MASS::mvrnorm(n1, rep(0, p), Sigma)
eta1 <- X1 %*% Theta_true + matrix(beta_true, n1, K - 1, byrow = TRUE)
P1   <- softmax_first(eta1)
y    <- apply(P1, 1, function(pp) sample.int(K, 1, prob = pp))

# qhat = P(Group 2 | X) = 1 - P(W=1 | X)
dnew <- xgb.DMatrix(data = X1[, phi_idx])
q1   <- as.numeric(predict(xgb_fit, newdata = dnew))
qhat <- matrix(1 - q1, ncol = 1)
```

## 4. Fit MLE and FMLE

```{r fit}
# --- Baseline MLE via nnet ---
df <- data.frame(y = factor(y), X1)
fit_mle <- nnet::multinom(y ~ ., data = df, trace = FALSE)
cf <- coef(fit_mle)
if (is.vector(cf)) cf <- matrix(cf, nrow = 1)

beta0  <- cf[, "(Intercept)"]
Theta0 <- t(cf[, colnames(cf) != "(Intercept)", drop = FALSE])

se_cf <- summary(fit_mle)$standard.errors
if (is.vector(se_cf)) se_cf <- matrix(se_cf, nrow = 1)
se_beta0  <- se_cf[, "(Intercept)"]
se_Theta0 <- t(se_cf[, colnames(se_cf) != "(Intercept)", drop = FALSE])

# --- FMLE (warm-start from MLE) ---
Hmat <- build_H(X1, phi_idx = phi_idx, n.q = nq)
t0   <- matrix(0, nrow = ncol(qhat), ncol = ncol(Hmat))
par0 <- pack_hard(beta0, Theta0, rep(0, K - 1), t0)

fit_fmle <- ml_fused(
  par = par0, X = X1, y = y,
  qhat = qhat, Hmat = Hmat, groups = groups,
 tau_tmat = 0.1,
  maxit = 500, tol = 1e-6
)
```

## 5. Coefficient plot

```{r coefplot, fig.width = 9, fig.height = 5}
z <- qnorm(0.975)

param_names <- c(
  paste0("beta[", 1:(K - 1), "]"),
  paste0("Theta[", rep(1:p, K - 1), ",", rep(1:(K - 1), each = p), "]")
)
truth <- c(beta_true, as.vector(Theta_true))

# MLE
est_mle <- c(beta0, as.vector(Theta0))
se_mle  <- c(se_beta0, as.vector(se_Theta0))

# FMLE
est_fmle <- c(fit_fmle$par$beta, as.vector(fit_fmle$par$Theta))
se_fmle  <- c(fit_fmle$se$beta, as.vector(fit_fmle$se$Theta))

plot_df <- data.frame(
  param    = rep(param_names, 2),
  method   = rep(c("MLE", "FMLE"), each = length(truth)),
  estimate = c(est_mle, est_fmle),
  se       = c(se_mle, se_fmle),
  truth    = rep(truth, 2)
)
plot_df$lwr   <- plot_df$estimate - z * plot_df$se
plot_df$upr   <- plot_df$estimate + z * plot_df$se
plot_df$param <- factor(plot_df$param, levels = param_names)

ggplot(plot_df, aes(x = param, y = estimate, color = method)) +
  geom_hline(aes(yintercept = truth), linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(ymin = lwr, ymax = upr),
                  position = position_dodge(width = 0.5), size = 0.5) +
  scale_color_manual(values = c(MLE = "#E69F00", FMLE = "#0072B2")) +
  labs(title = "MLE vs FMLE: Estimates with 95% CI",
       subtitle = "Dashed lines = true values",
       x = NULL, y = "Estimate", color = "Method") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")
```
