---
title: "NHANES Blood Pressure Application"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NHANES Blood Pressure Application}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 6
)
has_nnet    <- requireNamespace("nnet", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
has_deps    <- has_nnet && has_ggplot2
```

## Overview

This vignette reproduces the NHANES blood pressure analysis from the
accompanying paper. We fit multinomial logistic regression models for a
three-class blood pressure outcome (Normal, Prehypertension, Hypertension)
on primary (fasting) NHANES data, incorporating external ML predictions via
the MLfused empirical-likelihood framework.

The analysis demonstrates two settings:

1. **Coarsened external labels** -- the external ML model predicts a binary
   grouping: {Normal} vs {Prehypertension, Hypertension}.
2. **Full external labels** -- the external ML model predicts all three
   classes separately.

For each setting, we compare the standard multinomial logistic regression
with the fused estimator on two subsets of the primary data: the full
fasting subsample and a random subsample of 600 observations.

## Load data

The processed NHANES datasets are bundled in `inst/extdata`. These were
constructed from NHANES cycles H, I, and J using the pipeline in
`inst/scripts`.

```{r load-data, eval = has_deps}
library(MLfused)

# Paths to bundled data
path_coarsened <- system.file("extdata", "internal_pred_coarsened.csv",
                              package = "MLfused")
path_full      <- system.file("extdata", "internal_pred_full.csv",
                              package = "MLfused")

data_coarsened <- read.csv(path_coarsened, stringsAsFactors = FALSE)
data_full      <- read.csv(path_full, stringsAsFactors = FALSE)

cat("Coarsened data:", nrow(data_coarsened), "rows\n")
cat("Full data:     ", nrow(data_full), "rows\n")
```

## Prepare covariates

We use the same covariate set as the paper: age, sex, race, income-to-poverty
ratio, BMI, waist circumference, height, weight, and log-transformed lab
values (glucose, insulin, triglycerides, LDL, HDL, total cholesterol).
Continuous variables are centered and scaled.

```{r prepare-covariates, eval = has_deps}
var_conti <- c("age", "income_pir", "bmi", "waist_cm", "height_cm", "weight_kg")
var_outcome <- "bp_cat"
var_all <- c("age", "sex", "race", "income_pir", "bmi", "waist_cm",
             "height_cm", "weight_kg", "glucose_mg_dl_log1p",
             "insulin_uU_mL_log1p", "triglycerides_mg_dl_log1p",
             "ldl_mg_dl_log1p", "hdl_mg_dl_log1p", "total_chol_mg_dl_log1p")

prepare_data <- function(data) {
  # Factor variables
  data$sex  <- as.factor(data$sex)
  data$race <- as.factor(data$race)
  data$bp_cat <- factor(data$bp_cat,
                        levels = c("Normal", "Prehypertension", "Hypertension"))

  # Design matrix (numeric)
  X_raw    <- data.matrix(data[, var_all])
  X_scaled <- scale(X_raw)
  y        <- as.integer(data$bp_cat)

  # Continuous covariates (unscaled) for H matrix construction
  X_conti <- as.data.frame(X_raw[, var_conti, drop = FALSE])

  list(X = X_scaled, y = y, X_conti = X_conti,
       x_colnames = colnames(X_scaled), data = data)
}
```

## Helper functions

```{r helpers, eval = has_deps}
# Format coefficient rows into a data frame
make_coef_rows <- function(subset_name, label_type, method, beta, Theta,
                           x_colnames, se_mat = NULL, nq = NA) {
  K_minus1   <- length(beta)
  coef_names <- c("intercept", x_colnames)
  rows <- lapply(seq_len(K_minus1), function(ki) {
    est <- c(beta[ki], Theta[, ki])
    se  <- if (!is.null(se_mat)) se_mat[ki, ] else rep(NA_real_, length(coef_names))
    data.frame(
      subset_name = subset_name,
      label_type  = label_type,
      Method      = method,
      n_q         = nq,
      coef_name   = coef_names,
      class_idx   = ki + 1L,
      estimate    = est,
      std_error   = se,
      ci_lo       = est - 1.96 * se,
      ci_hi       = est + 1.96 * se,
      stringsAsFactors = FALSE
    )
  })
  do.call(rbind, rows)
}
```

```{r plot-helper, eval = has_deps}
# Comparison figure: two subsets, Multinomial vs one Fused method
make_comparison_figure <- function(res, subset_A, subset_B,
                                   label_A, label_B,
                                   fused_method, title_str) {
  library(ggplot2)

  coef_order <- c("intercept", "age", "sex", "race", "income_pir",
                  "bmi", "waist_cm", "height_cm", "weight_kg",
                  "glucose_mg_dl_log1p", "insulin_uU_mL_log1p",
                  "triglycerides_mg_dl_log1p", "ldl_mg_dl_log1p",
                  "hdl_mg_dl_log1p", "total_chol_mg_dl_log1p")
  coef_short <- c(
    intercept = "(Int)", age = "Age", sex = "Sex", race = "Race",
    income_pir = "Income", bmi = "BMI", waist_cm = "Waist",
    height_cm = "Height", weight_kg = "Weight",
    glucose_mg_dl_log1p = "Glucose", insulin_uU_mL_log1p = "Insulin",
    triglycerides_mg_dl_log1p = "TG", ldl_mg_dl_log1p = "LDL",
    hdl_mg_dl_log1p = "HDL", total_chol_mg_dl_log1p = "Total Chol")
  class_labels <- c(`2` = "Prehypertension vs Normal",
                    `3` = "Hypertension vs Normal")

  plot_dat <- res[res$subset_name %in% c(subset_A, subset_B) &
                  res$Method %in% c("Multinomial", fused_method), ]

  # Map subset names to display labels
  plot_dat$subset_label <- ifelse(plot_dat$subset_name == subset_A, label_A, label_B)
  plot_dat$subset_label <- factor(plot_dat$subset_label, levels = c(label_A, label_B))
  plot_dat$method_label <- factor(plot_dat$Method,
                                  levels = c("Multinomial", fused_method))
  plot_dat$coef_name    <- factor(plot_dat$coef_name, levels = coef_order)
  plot_dat$coef_display <- factor(coef_short[as.character(plot_dat$coef_name)],
                                  levels = coef_short[coef_order])
  plot_dat$class_label  <- factor(class_labels[as.character(plot_dat$class_idx)],
                                  levels = class_labels)
  plot_dat <- plot_dat[!is.na(plot_dat$coef_name) & !is.na(plot_dat$class_label), ]

  if (nrow(plot_dat) == 0) return(invisible(NULL))

  method_colors <- c("Multinomial" = "steelblue", setNames("tomato3", fused_method))
  method_shapes <- c("Multinomial" = 16, setNames(15, fused_method))

  ggplot(plot_dat, aes(x = coef_display, y = estimate,
                       shape = method_label, color = method_label)) +
    geom_hline(yintercept = 0, color = "gray80", linewidth = 0.4) +
    geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                  position = position_dodge(width = 0.6),
                  width = 0.3, linewidth = 0.4, alpha = 0.75) +
    geom_point(size = 1.8, position = position_dodge(width = 0.6)) +
    facet_grid(class_label ~ subset_label) +
    scale_shape_manual(name = "Method", values = method_shapes) +
    scale_color_manual(name = "Method", values = method_colors) +
    theme_bw(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8.5),
          legend.position = "bottom",
          strip.text = element_text(face = "bold"),
          panel.grid.minor = element_blank()) +
    labs(x = NULL, y = "Coefficient estimate",
         title = title_str,
         subtitle = "Error bars: estimate +/- 1.96 x SE")
}
```

## Analysis: Coarsened external labels

In the coarsened setting, the external ML model predicts a binary outcome:
{Normal} vs {Prehypertension + Hypertension}. This gives `L = 2` groups with
group 1 = {Normal} as the reference.

```{r coarsened-analysis, eval = has_deps}
# Prepare data
d <- prepare_data(data_coarsened)
X_scaled   <- d$X
y          <- d$y
X_conti    <- d$X_conti
x_colnames <- d$x_colnames
N          <- nrow(X_scaled)

# External predictions: coarsened (binary)
qhat_c    <- matrix(data_coarsened$qhat,    ncol = 1)
qhat_se_c <- matrix(data_coarsened$qhat_se, ncol = 1)

# Groups: {Normal} vs {Prehypertension, Hypertension}
groups_c <- list(c(1), c(2, 3))
Lm1_c    <- length(groups_c) - 1  # = 1

# Define subsets
set.seed(42)
subset_defs <- list(
  fasting_all = seq_len(N),
  random_600  = sort(sample(seq_len(N), 600, replace = FALSE))
)

# Tuning parameters
LAMBDA <- nrow(data_coarsened)
NQ     <- 0  # intercept + centered covariates

results_c <- data.frame()

for (subset_name in names(subset_defs)) {
  idx <- subset_defs[[subset_name]]
  X_sub       <- X_scaled[idx, , drop = FALSE]
  y_sub       <- y[idx]
  qhat_sub    <- qhat_c[idx, , drop = FALSE]
  qhat_se_sub <- qhat_se_c[idx, , drop = FALSE]
  X_conti_sub <- X_conti[idx, , drop = FALSE]
  n_sub       <- length(idx)

  cat("\n--- Subset:", subset_name, "(n =", n_sub, ") ---\n")

  # Multinomial baseline
  fit_mn <- nnet::multinom(y_sub ~ X_sub, trace = FALSE)
  cf_mn  <- coef(fit_mn)
  if (is.vector(cf_mn)) cf_mn <- matrix(cf_mn, nrow = 1)
  beta_mn  <- cf_mn[, "(Intercept)"]
  Theta_mn <- t(cf_mn[, colnames(cf_mn) != "(Intercept)", drop = FALSE])
  se_mn    <- summary(fit_mn)$standard.errors
  se_mn_mat <- cbind(
    se_mn[, "(Intercept)", drop = FALSE],
    se_mn[, paste0("X_sub", x_colnames), drop = FALSE]
  )

  results_c <- rbind(results_c,
    make_coef_rows(subset_name, "coarsened", "Multinomial",
                   beta_mn, Theta_mn, x_colnames, se_mat = se_mn_mat))

  # Fused model
  H_base   <- build_H(X_conti_sub, phi_idx = seq_along(var_conti), n.q = NQ)
  Hmat_sub <- build_Hmat_interleaved(H_base, Lm1_c)

  alpha0 <- rep(0, ncol(Theta_mn))
  tmat0  <- matrix(0, nrow = Lm1_c, ncol = ncol(H_base))
  par0   <- pack.hard(beta_mn, Theta_mn, alpha0, tmat0)

  fit_f <- tryCatch(
    ML_fused_fast(
      par           = par0,
      X             = X_sub,
      y             = y_sub,
      qhat          = qhat_sub,
      qhat.se       = qhat_se_sub,
      Hmat          = Hmat_sub,
      groups        = groups_c,
      maxit         = 500,
      tol           = 1e-6,
      learning.rate = 0.1,
      lambda        = LAMBDA,
      lambda.diag   = 0,
      compute_se    = TRUE,
      tau_tmat      = 0.01
    ),
    error = function(e) { message("  Fused ERROR: ", e$message); NULL }
  )

  if (!is.null(fit_f)) {
    cat("  Convergence:", fit_f$conv, "\n")
    se_score <- fit_f$se.score.adj
    if (!is.null(se_score)) {
      se_score_mat <- cbind(se_score$beta, t(se_score$Theta))
    } else {
      se_score_mat <- NULL
    }
    results_c <- rbind(results_c,
      make_coef_rows(subset_name, "coarsened", "Fused (n.q=0)",
                     fit_f$par$beta, fit_f$par$Theta,
                     x_colnames, se_mat = se_score_mat, nq = NQ))
  }
}
```

### Coefficient comparison figure (coarsened)

```{r coarsened-figure, fig.width = 10, fig.height = 6.5, eval = has_deps}
p_coarsened <- make_comparison_figure(
  results_c,
  subset_A = "fasting_all", subset_B = "random_600",
  label_A  = "Fasting (All)", label_B = "Random 600",
  fused_method = "Fused (n.q=0)",
  title_str = "[Coarsened] Fasting All vs Random 600"
)
if (!is.null(p_coarsened)) print(p_coarsened)
```

## Analysis: Full external labels

In the full setting, the external ML model predicts all three classes
separately. This gives `L = 3` groups with group 1 = {Normal} as the
reference.

```{r full-analysis, eval = has_deps}
d2 <- prepare_data(data_full)
X_scaled2   <- d2$X
y2          <- d2$y
X_conti2    <- d2$X_conti
x_colnames2 <- d2$x_colnames
N2          <- nrow(X_scaled2)

# External predictions: full (3-class)
qhat_f    <- as.matrix(data_full[, c("qhat_prehyp", "qhat_hyp")])
qhat_se_f <- as.matrix(data_full[, c("qhat_se_prehyp", "qhat_se_hyp")])

# Groups: {Normal}, {Prehypertension}, {Hypertension}
groups_f <- list(c(1), c(2), c(3))
Lm1_f    <- length(groups_f) - 1  # = 2

set.seed(42)
subset_defs2 <- list(
  fasting_all = seq_len(N2),
  random_600  = sort(sample(seq_len(N2), 600, replace = FALSE))
)

results_f <- data.frame()

for (subset_name in names(subset_defs2)) {
  idx <- subset_defs2[[subset_name]]
  X_sub       <- X_scaled2[idx, , drop = FALSE]
  y_sub       <- y2[idx]
  qhat_sub    <- qhat_f[idx, , drop = FALSE]
  qhat_se_sub <- qhat_se_f[idx, , drop = FALSE]
  X_conti_sub <- X_conti2[idx, , drop = FALSE]
  n_sub       <- length(idx)

  cat("\n--- Subset:", subset_name, "(n =", n_sub, ") ---\n")

  # Multinomial baseline
  fit_mn <- nnet::multinom(y_sub ~ X_sub, trace = FALSE)
  cf_mn  <- coef(fit_mn)
  if (is.vector(cf_mn)) cf_mn <- matrix(cf_mn, nrow = 1)
  beta_mn  <- cf_mn[, "(Intercept)"]
  Theta_mn <- t(cf_mn[, colnames(cf_mn) != "(Intercept)", drop = FALSE])
  se_mn    <- summary(fit_mn)$standard.errors
  se_mn_mat <- cbind(
    se_mn[, "(Intercept)", drop = FALSE],
    se_mn[, paste0("X_sub", x_colnames2), drop = FALSE]
  )

  results_f <- rbind(results_f,
    make_coef_rows(subset_name, "full", "Multinomial",
                   beta_mn, Theta_mn, x_colnames2, se_mat = se_mn_mat))

  # Fused model with interleaved Hmat for L=3
  H_base   <- build_H(X_conti_sub, phi_idx = seq_along(var_conti), n.q = NQ)
  Hmat_sub <- build_Hmat_interleaved(H_base, Lm1_f)

  alpha0 <- rep(0, ncol(Theta_mn))
  tmat0  <- matrix(0, nrow = Lm1_f, ncol = ncol(H_base))
  par0   <- pack.hard(beta_mn, Theta_mn, alpha0, tmat0)

  fit_fused <- tryCatch(
    ML_fused_fast(
      par           = par0,
      X             = X_sub,
      y             = y_sub,
      qhat          = qhat_sub,
      qhat.se       = qhat_se_sub,
      Hmat          = Hmat_sub,
      groups        = groups_f,
      maxit         = 500,
      tol           = 1e-6,
      learning.rate = 0.1,
      lambda        = LAMBDA,
      lambda.diag   = 0,
      compute_se    = TRUE,
      tau_tmat      = 0.01
    ),
    error = function(e) { message("  Fused ERROR: ", e$message); NULL }
  )

  if (!is.null(fit_fused)) {
    cat("  Convergence:", fit_fused$conv, "\n")
    se_score <- fit_fused$se.score.adj
    if (!is.null(se_score)) {
      se_score_mat <- cbind(se_score$beta, t(se_score$Theta))
    } else {
      se_score_mat <- NULL
    }
    results_f <- rbind(results_f,
      make_coef_rows(subset_name, "full", "Fused (n.q=0)",
                     fit_fused$par$beta, fit_fused$par$Theta,
                     x_colnames2, se_mat = se_score_mat, nq = NQ))
  }
}
```

### Coefficient comparison figure (full)

```{r full-figure, fig.width = 10, fig.height = 6.5, eval = has_deps}
p_full <- make_comparison_figure(
  results_f,
  subset_A = "fasting_all", subset_B = "random_600",
  label_A  = "Fasting (All)", label_B = "Random 600",
  fused_method = "Fused (n.q=0)",
  title_str = "[Full 3-class] Fasting All vs Random 600"
)
if (!is.null(p_full)) print(p_full)
```

## Interpreting the results

Several patterns are evident in the figures:

- **Coefficient estimates** from the fused model are generally close to
  those of the standard multinomial, confirming that the external information
  does not introduce substantial bias.
- **Confidence intervals** for the fused estimator tend to be narrower,
  reflecting the efficiency gain from incorporating external predictions.
- The benefit is more pronounced for the **random 600 subset**, where the
  smaller sample size makes the external information relatively more valuable.
- In the coarsened setting (binary external labels), the fused model still
  improves estimation of the full 3-class multinomial, demonstrating that
  even partial external information is useful.

## Session info

```{r session-info}
sessionInfo()
```
